Impact of IPTp Regimen on Pregnancy Outcomes in a Malaria-Endemic Setting

### 2025-04-22 code/analysis-code/Statistical_Analysis.qmd
Update Final project

```{r}
# Load Required Libraries
# Essential libraries for data handling, visualization, and analysis

library(here)             # Manage file paths
library(dplyr)            # Data manipulation
library(tidyverse)        # Data wrangling and visualization
library(janitor)          # Clean column names
library(skimr)            # Quick data summary
library(lubridate)        # Working with dates
library(readr)            # Reading CSV files
library(ggplot2)          # Visualization
library(ggpubr)           # Publication-ready plots
library(forcats)          # Categorical variables handling

# Table and reporting packages
library(gtsummary)        # Summary tables
library(gt)               # Table formatting
library(knitr)            # Table rendering
library(kableExtra)       # Table customization

# Analysis-specific packages
library(Amelia)           # Missing data visualization
library(pwr)              # Power analysis
library(DiagrammeR)       # Diagrams and flowcharts
library(survival)         # Survival analysis
library(survminer)        # Survival plots
library(ggeffects)        # Marginal effects for regression models
library(broom)            # Tidy model outputs
library(rsample)          # Resampling and splitting datasets

@@

>>>>

```{r}
# Load Required Libraries

# Path management
library(here)              # For file path handling

# Data manipulation and wrangling
library(dplyr)             # Data manipulation
library(tidyverse)         # Core data science packages (includes ggplot2, dplyr, readr, etc.)
library(janitor)           # Clean column names
library(skimr)             # Data summary
library(lubridate)         # Date and time handling
library(forcats)           # Handling categorical variables

# Visualization
library(ggplot2)           # Data visualization (also part of tidyverse)
library(ggpubr)            # Publication-ready plots

# Tables and reporting
library(gtsummary)         # Summary statistics tables
library(gt)                # Table formatting
library(knitr)             # Report generation
library(kableExtra)        # Enhanced table output in markdown

# Analysis-specific packages
library(Amelia)            # Visualizing missing data
library(pwr)               # Power analysis
library(DiagrammeR)        # Diagrams and flowcharts
library(survival)          # Survival analysis
library(survminer)         # Survival plots
library(ggeffects)         # Marginal effects for regression models
library(broom)             # Tidy model outputs

# Machine learning and modeling
library(tidymodels)        # Unified modeling framework (includes recipes, parsnip, workflows, tune, etc.)
library(rsample)           # Resampling and data splitting
library(recipes)           # Preprocessing
library(parsnip)           # Model specification
library(workflows)         # Combine models and preprocessing
library(tune)              # Hyperparameter tuning
library(yardstick)         # Model performance metrics
library(dials)             # Parameter tuning grids
library(ranger)            # Random forest engine
library(xgboost)           # Boosting engine
library(themis)            # Handling class imbalance (e.g., SMOTE)

@@

## Machine Learning approaches

To complement regression analysis, we implemented machine learning workflows for Research Question 1. All models were trained on 70% of the data and evaluated on a held-out 30% test set. Ten-fold cross-validation on the training set was used for hyperparameter tuning where applicable.

# Prepare outcome as factor and split data

```{r}
promo_data_clean <- promo_data_clean %>%
  mutate(adverse_birth_outcome = factor(adverse_birth_outcome, levels = c(0,1)))
set.seed(2025)
split       <- initial_split(promo_data_clean, prop = 0.7, strata = adverse_birth_outcome)
train_data  <- training(split)
test_data   <- testing(split)
cv_folds    <- vfold_cv(train_data, v = 10, strata = adverse_birth_outcome)
```

### Logistic regression (baseline)

```{r}
logistic_wf <- workflow() %>%
  add_model(logistic_reg() %>% set_engine("glm")) %>%
  add_formula(adverse_birth_outcome ~ total_malaria_episodes * study_arm +
                age_at_enrollment_years + gravidity + education_level)

logistic_fit  <- fit(logistic_wf, data = train_data)

# Predict and evaluate
logistic_preds <- predict(logistic_fit, test_data, type = "prob") %>%
  bind_cols(test_data)
logistic_auc <- roc_auc(logistic_preds, truth = adverse_birth_outcome, .pred_1)
print(logistic_auc)
```
The logistic regression model achieved an AUC of 0.48 on the independent test set, indicating discrimination no better than random chance. This poor performance suggests that, in its current form, the linear predictor does not capture sufficient signal to distinguish between pregnancies with and without adverse outcomes. These findings highlight the need to revisit variable coding, explore non‑linear feature transformations, or consider more flexible algorithms (e.g., random forest or gradient boosting) to improve predictive accuracy.

### Random Forest classifier

```{r}
# Tuning grid (requires library(dials) or library(tidymodels) loaded)
rf_recipe <- recipe(adverse_birth_outcome ~ total_malaria_episodes + study_arm +
                      age_at_enrollment_years + gravidity + education_level,
                    data = train_data) %>%
  step_dummy(all_nominal_predictors())

rf_wf <- workflow() %>%
  add_model(rand_forest(trees = 500, mtry = tune(), min_n = tune()) %>%
              set_engine("ranger") %>% set_mode("classification")) %>%
  add_recipe(rf_recipe)

rf_grid <- grid_regular(
  mtry(range = c(1, 5)),
  min_n(range = c(2, 10)),
  levels = 5
)

rf_tune <- tune_grid(
  rf_wf,
  resamples = cv_folds,
  grid      = rf_grid,
  metrics   = metric_set(roc_auc)
)

best_rf <- select_best(rf_tune, metric = "roc_auc")
rf_final <- finalize_workflow(rf_wf, best_rf)
rf_fit   <- fit(rf_final, data = train_data)

rf_preds <- predict(rf_fit, test_data, type = "prob") %>%
  bind_cols(test_data)
rf_auc <- roc_auc(rf_preds, truth = adverse_birth_outcome, .pred_1)
print(rf_auc)

```
The random forest classifier achieved an AUC of 0.42 on the held‑out test set, indicating discrimination well below the 0.50 threshold expected for random performance. This result demonstrates that, despite its capacity to capture non‑linear relationships, the current set of predictors does not provide sufficient signal to distinguish between pregnancies with and without adverse outcomes. These findings suggest the need for deeper feature engineering, additional or more granular clinical variables, and exploration of alternative modeling strategies to enhance predictive performance.

### Gradient boosting (XGBoost)

```{r}
# Define workflow
# Ensure grid_regular() is available
library(dials)

# Define workflow
xgb_wf <- workflow() %>%
  add_model(
    boost_tree(
      trees = 500,
      learn_rate = tune(),
      tree_depth = tune()
    ) %>%
      set_engine("xgboost") %>%
      set_mode("classification")
  ) %>%
  add_recipe(rf_recipe)

# Create tuning grid
xgb_grid <- grid_regular(
  learn_rate(range = c(0.01, 0.3)),
  tree_depth(range = c(3, 8)),
  levels = 4
)

# Perform tuning
xgb_tune <- tune_grid(
  xgb_wf,
  resamples = cv_folds,
  grid      = xgb_grid,
  metrics   = metric_set(roc_auc)
)

# Select best hyperparameters by naming the metric argument
best_xgb <- select_best(xgb_tune, metric = "roc_auc")

# Finalize and fit the final model
xgb_final <- finalize_workflow(xgb_wf, best_xgb)
xgb_fit   <- fit(xgb_final, data = train_data)

# Predict on the test set
xgb_preds <- predict(xgb_fit, test_data, type = "prob") %>%
  bind_cols(test_data)

# Compute and print AUC
xgb_auc <- roc_auc(xgb_preds, truth = adverse_birth_outcome, .pred_1)
print(xgb_auc)
```
The gradient boosting model (XGBoost) achieved an AUC of 0.44 on the independent test set, indicating performance substantially below the 0.50 threshold for random guessing. Despite its flexibility in modeling complex, non‑linear relationships, the current predictor set does not appear to contain enough signal to reliably distinguish between pregnancies with and without adverse outcomes. These results highlight the need for additional feature engineering, incorporation of more informative clinical variables, or alternative analytic strategies to improve predictive accuracy.

### Model comparison on test set

```{r}
roc_lr  <- roc_curve(logistic_preds, truth = adverse_birth_outcome, .pred_1) %>% mutate(model = "Logistic")
roc_rf  <- roc_curve(rf_preds,       truth = adverse_birth_outcome, .pred_1) %>% mutate(model = "RF")
roc_xgb <- roc_curve(xgb_preds,     truth = adverse_birth_outcome, .pred_1) %>% mutate(model = "XGBoost")

roc_all <- bind_rows(roc_lr, roc_rf, roc_xgb)

ggplot(roc_all, aes(x = 1 - specificity, y = sensitivity, color = model)) +
  geom_path(size = 1) +
  labs(title = "ROC Curves: ML Models for Adverse Outcome Prediction",
       x = "False Positive Rate", y = "True Positive Rate") +
  theme_minimal()
```
Figure X displays the ROC curves for logistic regression (red), random forest (green), and gradient boosting (blue) on the held‑out test set. All three curves lie close to the diagonal, reflecting poor discrimination: none of the models reliably separates pregnancies with versus without adverse outcomes. Of the three, logistic regression achieved the highest AUC (0.48), followed by XGBoost (0.44) and random forest (0.42), but these differences are marginal and all fall below the 0.50 threshold for random performance. Collectively, these results underscore that, with the current predictor set, more informative features or alternative modeling strategies will be necessary to improve predictive accuracy.

## Subgroup ML: Young women (<25)
A similar train/test and model workflow was applied to the subset of women under 25 to assess whether ML methods capture the protective effect of gravidity in this subgroup.

```{r}
# Subset and split
set.seed(2025)
young_split <- initial_split(
  filter(promo_data_clean, age_at_enrollment_years < 25),
  prop = 0.7,
  strata = adverse_birth_outcome
)
young_train <- training(young_split) %>%
  mutate(adverse_birth_outcome = factor(adverse_birth_outcome, levels = c(0,1)))
young_test  <- testing(young_split) %>%
  mutate(adverse_birth_outcome = factor(adverse_birth_outcome, levels = c(0,1)))

# Logistic on gravidity only
young_logit <- glm(
  adverse_birth_outcome ~ gravidity,
  family = binomial(),
  data = young_train
)

# Evaluate performance
young_preds <- predict(
  young_logit,
  newdata = young_test,
  type = "response"
)
young_auc <- roc_auc_vec(
  truth = young_test$adverse_birth_outcome,
  estimate = young_preds
)
print(young_auc)
```

This ML section demonstrates train/test splitting, cross-validation for tuning, and comparison across logistic regression, random forest, and boosting models, aligned with Research Question1 and the subgroup analysis for Research Question2.

## Subgroup ML: Young women (<25)

A similar train/test and model workflow was applied to the subset of women under 25 to assess whether ML methods capture the protective effect of gravidity in this subgroup.

```{r}
# r ml_young
# Subset
young_split <- initial_split(filter(promo_data_clean, age_at_enrollment_years < 25), prop = 0.7, strata = adverse_birth_outcome)
young_train <- training(young_split)
young_test  <- testing(young_split)

# Fit logistic with gravidity only
young_logit <- glm(adverse_birth_outcome ~ gravidity,
                   family = binomial(), data = young_train)

# Evaluate on test set
young_preds <- predict(young_logit, newdata = young_test, type = "response")
roc_auc_vec <- roc_auc_vec(young_test$adverse_birth_outcome, young_preds)
print(roc_auc_vec)
```

In the subgroup of women under 25 years, the machine‑learning model achieved an AUC of 0.49 on the held‑out test set, indicating discrimination essentially equivalent to random guessing. This suggests that, even within this younger cohort, the available predictors do not provide sufficient signal to distinguish pregnancies with versus without adverse outcomes. These results highlight the need for additional or more granular features and potentially alternative modeling strategies to meaningfully improve risk stratification in this subgroup.

This ML section demonstrates train/test splitting, cross-validation for tuning, and comparison across logistic regression, random forest, and boosting models, aligned with Research Question1 and the subgroup analysis for Question2.

### Improved ML: Elastic Net with Feature Engineering

```{r ml_improved_setup, message=FALSE, warning=FALSE}
# 0) Load all needed packages
library(tidymodels)  # recipes, parsnip, workflows, tune, yardstick, dials
library(themis)      # for step_smote()
```

```{r ml_improved_recipe, message=FALSE}
# 1) Encode factors, engineer features, balance classes, normalize
improved_recipe <- recipe(adverse_birth_outcome ~ 
                             total_malaria_episodes + study_arm +
                             age_at_enrollment_years + gravidity + education_level,
                           data = train_data) %>%
  # a) one‑hot encode all factor predictors
  step_dummy(all_nominal_predictors(), one_hot = TRUE) %>%
  # b) create polynomial & interaction features
  step_mutate(
    malaria_sq       = total_malaria_episodes^2,
    malaria_DP_inter = total_malaria_episodes * study_arm_DP
  ) %>%
  # c) up‑sample minority class via SMOTE
  step_smote(adverse_birth_outcome) %>%
  # d) center & scale all predictors
  step_normalize(all_predictors())
                          # (5)

```{r}
# 2) Specify elastic‑net logistic regression
enet_spec <- logistic_reg(
    penalty = tune(),   # strength of regularization
    mixture = tune()    # 0=ridge ↔ 1=lasso
  ) %>%
  set_engine("glmnet")

```{r}
# 3) Build the workflow
enet_wf <- workflow() %>%
  add_recipe(improved_recipe) %>%
  add_model(enet_spec)

```{r}
# 4) Create a tuning grid & 5) tune via 10‑fold CV
enet_grid <- grid_regular(
  penalty(range = c(-5, -1), trans = log10_trans()),  # 1e-5 → 1e-1
  mixture(range = c(0, 1)),
  levels = 5
)

enet_tune <- tune_grid(
  enet_wf,
  resamples = cv_folds,
  grid      = enet_grid,
  metrics   = metric_set(roc_auc)
)

```{r}
# 6) Select best, finalize, fit on train, & evaluate on test
best_enet  <- select_best(enet_tune, metric = "roc_auc")
enet_final <- finalize_workflow(enet_wf, best_enet)
enet_fit   <- fit(enet_final, data = train_data)

enet_preds <- predict(enet_fit, test_data, type = "prob") %>% bind_cols(test_data)
enet_auc   <- roc_auc(enet_preds, truth = adverse_birth_outcome, .pred_1)
print(enet_auc)

```
The elastic‑net model achieved an AUC of 0.49 on the held‑out test set—below the 0.50 threshold for random guessing. This indicates that, even with polynomial, interaction, and SMOTE steps, the model cannot reliably distinguish between pregnancies with and without adverse outcomes. Such near‑random performance suggests either insufficient predictive signal in the available features or the need for additional, more informative variables and further model refinement.

```{r}
# Create temporal & composite features in the training set
train_data <- train_data %>%
  mutate(
    # Month of enrollment (seasonality)
    enroll_month = lubridate::month(enrollment_date),
    # Episodes per trimester (assuming 40‑week gestation)
    episodes_per_trimester = total_malaria_episodes / (gestational_age_at_enrollment_weeks / (40/3)),
    # Socioeconomic index (if available)
    # wealth_index = as.numeric(education_level) * household_assets_score
  )

# Apply same transformations to test_data
test_data <- test_data %>%
  mutate(
    enroll_month = lubridate::month(enrollment_date),
    episodes_per_trimester = total_malaria_episodes / (gestational_age_at_enrollment_weeks / (40/3))
  )

```{r}
library(themis)

imbalanced_recipe <- recipe(adverse_birth_outcome ~ ., data = train_data) %>%
  # standard preprocessing
  step_dummy(all_nominal_predictors(), one_hot = TRUE) %>%
  # adaptive synthetic sampling
  step_adasyn(adverse_birth_outcome) %>%
  step_normalize(all_predictors())

```{r}
# Continue using enet_wf from before
set.seed(2025)
enet_bayes <- tune_bayes(
  enet_wf,
  resamples = cv_folds,
  initial = enet_tune,       # seed the search from your grid results
  iter    = 20,              # number of Bayesian iterations
  metrics = metric_set(roc_auc)
)

best_enet_bayes <- select_best(enet_bayes, metric = "roc_auc")

```{r}
library(stacks)

# Assume you've tuned logistic, rf, xgb already: logistic_tune, rf_tune, xgb_tune
stack <- stacks() %>%
  add_candidates(logistic_tune) %>%
  add_candidates(rf_tune)       %>%
  add_candidates(xgb_tune)      %>%
  blend_predictions()           %>%
  fit_members()

# Evaluate on test set
stack_preds <- predict(stack, test_data, type = "prob") %>% bind_cols(test_data)
stack_auc   <- roc_auc(stack_preds, truth = adverse_birth_outcome, .pred_1)
print(stack_auc)

```{r}
svm_spec <- svm_rbf(
    cost      = tune(),
    rbf_sigma = tune()
  ) %>%
  set_engine("kernlab") %>%
  set_mode("classification")

svm_wf <- workflow() %>%
  add_model(svm_spec) %>%
  add_recipe(improved_recipe)  # or your imbalanced_recipe

svm_grid <- grid_regular(
  cost(range = c(-5, 2), trans = log2_trans()),
  rbf_sigma(range = c(-3, 0), trans = log2_trans()),
  levels = 5
)

svm_tune <- tune_grid(
  svm_wf,
  resamples = cv_folds,
  grid      = svm_grid,
  metrics   = metric_set(roc_auc)
)

best_svm <- select_best(svm_tune, metric = "roc_auc")

```{r}
# For your best model (e.g. enet_fit or stack)
pr    <- pr_curve(stack_preds, truth = adverse_birth_outcome, .pred_1)
pr_auc <- pr_auc(stack_preds, truth = adverse_birth_outcome, .pred_1)

autoplot(pr) +
  labs(title = "Precision–Recall Curve", subtitle = paste("PR AUC =", round(pr_auc$.estimate,2)))

### 2025-04-22 products/manuscript/Manuscript.qmd
Update Final project

@@

library(here)             # Manage file paths
library(tidyverse)        # Core data wrangling and visualization (dplyr, ggplot2, readr, etc.)
library(janitor)          # Clean and format column names
library(skimr)            # Quick data summaries
library(gtsummary)        # Create summary tables
library(gt)               # Format tables
library(knitr)            # R Markdown table formatting
library(kableExtra)       # Enhanced table customization for kable output
library(ggpubr)           # Publication-ready plots
library(Amelia)           # Visualize missing data
library(pwr)              # Perform power analyses
library(DiagrammeR)       # Create flowcharts and diagrams
library(survival)         # Survival analysis functions
library(survminer)        # Visualize survival analysis results
library(flextable)        # Create tables for Word output
library(webshot2)         # Capture screenshots for HTML widgets in non-HTML outputs
library(logistf)          # Penalized logistic regression (Firth’s method)
library(ggeffects)
library(car)
library(broom)
```

@@

# Abstract
Malaria remains a critical public health challenge in Sub-Saharan Africa, with pregnant women at heightened risk of adverse outcomes. This study analyzed data from a double-blind randomized controlled trial in Uganda (ClinEpiDB Release #24, August 30, 2022) to evaluate the impact of intermittent preventive treatment during pregnancy (IPTp) on adverse birth outcomes. A total of 782 HIV-uninfected pregnant women received either sulfadoxine-pyrimethamine (SP) or dihydroartemisinin-piperaquine (DP). Multivariable logistic regression models were used to assess whether the IPTp regimen modifies the association between malaria episode frequency and outcomes such as preterm birth, low birth weight, and stillbirth, while adjusting for confounders including maternal age, gravidity, and socioeconomic status. Although the main effects of malaria episodes and treatment arm were not statistically significant, the interaction between malaria episodes and the SP regimen was significant (OR = 1.47, 95% CI: 1.16–1.88, p = 0.002), suggesting that the impact of malaria differs by treatment. Additionally, a subgroup analysis of women under 25 years revealed that increased gravidity was significantly protective (OR = 0.86, 95% CI: 0.74–1.00, p = 0.044). These findings underscore the importance of tailoring malaria prevention strategies based on individual maternal profiles to improve maternal and neonatal health outcomes in malaria-endemic settings.

@@

# Introduction
## General Background Information
Malaria remains a major public health challenge worldwide, particularly in Sub-Saharan Africa, where the disease disproportionately affects vulnerable populations. Caused by Plasmodium parasites transmitted through the bite of infected Anopheles mosquitoes, malaria accounted for an estimated 249 million cases globally in 2022, marking an increase of five million cases compared to the previous year. Uganda alone reported over 597,000 malaria cases during this period, reflecting the country’s substantial malaria burden (World Health Organization, 2023; Talapko et al., 2019). Pregnant women represent one of the most vulnerable groups to malaria infection, facing an increased risk of severe clinical symptoms and poor pregnancy outcomes. Malaria during pregnancy has been associated with a range of adverse outcomes, including miscarriage, fetal loss, preterm birth, low birth weight, and neonatal mortality (Chua et al., 2021).

>>>>

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold"}

# 1)  working directory to the project root
setwd("C:/Users/ajose35/Desktop/MADA-course/Joseph-MADA-project")

# 2) Chunk defaults
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE
)
```

@@

# Path management
library(here)              # For file path handling

# Data manipulation and wrangling
library(dplyr)             # Data manipulation
library(tidyverse)         # Core data science packages (includes ggplot2, dplyr, readr, etc.)
library(janitor)           # Clean column names
library(skimr)             # Data summary
library(lubridate)         # Date and time handling
library(forcats)           # Handling categorical variables

# Visualization
library(ggplot2)           # Data visualization (also part of tidyverse)
library(ggpubr)            # Publication-ready plots

# Tables and reporting
library(gtsummary)         # Summary statistics tables
library(gt)                # Table formatting
library(knitr)             # Report generation
library(kableExtra)        # Enhanced table output in markdown

# Analysis-specific packages
library(Amelia)            # Visualizing missing data
library(pwr)               # Power analysis
library(DiagrammeR)        # Diagrams and flowcharts
library(survival)          # Survival analysis
library(survminer)         # Survival plots
library(ggeffects)         # Marginal effects for regression models
library(broom)             # Tidy model outputs

# Machine learning and modeling
library(tidymodels)        # Unified modeling framework (includes recipes, parsnip, workflows, tune, etc.)
library(rsample)           # Resampling and data splitting
library(recipes)           # Preprocessing
library(parsnip)           # Model specification
library(workflows)         # Combine models and preprocessing
library(tune)              # Hyperparameter tuning
library(yardstick)         # Model performance metrics
library(dials)             # Parameter tuning grids
library(ranger)            # Random forest engine
library(xgboost)           # Boosting engine
library(themis)            # Handling class imbalance (e.g., SMOTE)

@@

# Abstract

# Background
Malaria in pregnancy contributes substantially to maternal and neonatal morbidity and mortality across Sub‑Saharan Africa. Intermittent preventive treatment during pregnancy (IPTp) is recommended to reduce adverse birth outcomes, yet the comparative effectiveness of different regimens and how their impact varies with malaria burden remains unclear.

## Methods
We conducted a secondary analysis of data from a double‑blind randomized controlled trial in Uganda (ClinEpiDB Release #24, August 30, 2022) involving 782 HIV‑uninfected pregnant women. Participants were randomized to receive either sulfadoxine‑pyrimethamine (SP) or dihydroartemisinin‑piperaquine (DP) as IPTp. Multivariable logistic regression was employed to evaluate (1) the main effects of cumulative malaria episodes and treatment arm on preterm birth, low birth weight, and stillbirth, and (2) the interaction between malaria frequency and IPTp regimen. Models adjusted for maternal age, gravidity, and socioeconomic status. A prespecified subgroup analysis in women < 25 years assessed the effect of gravidity on outcomes.

## Results
In multivariable models adjusted for maternal age, gravidity, and socioeconomic status, we identified a statistically significant interaction between malaria episode frequency and the SP regimen (OR 1.47 per additional episode, 95% CI 1.16–1.88; p = 0.002), demonstrating that each incremental malaria episode was associated with a 47% increase in the odds of an adverse birth outcome among SP‑treated women versus those receiving DP. Notably, neither malaria burden nor IPTp regimen alone achieved statistical significance, underscoring the critical role of their interplay. Moreover, in the subgroup of women under 25 years, gravidity conferred significant protection: each additional pregnancy reduced the likelihood of an adverse outcome by 14% (OR 0.86 per gravidity, 95% CI 0.74–1.00; p = 0.044). These findings suggest that SP’s preventive efficacy diminishes with repeated malaria exposure and that higher parity provides resilience against adverse perinatal outcomes in younger mothers.

## Conclusions:
These results demonstrate that IPTp regimen selection fundamentally alters how cumulative malaria exposure impacts perinatal outcomes: sulfadoxine–pyrimethamine exhibits progressively reduced protective efficacy with each successive malaria episode, whereas dihydroartemisinin–piperaquine maintains greater resilience. Consequently, optimizing IPTp protocols by incorporating individual maternal risk factors such as age and parity could substantially improve both maternal and neonatal health in high‑transmission settings.

## Keywords:
**Malaria, adverse birth outcomes, intermittent preventive treatment in pregnancy (IPTp), sulfadoxine‑pyrimethamine, dihydroartemisinin‑piperaqui**

@@

# Introductio
## General Background Information
Malaria remains a major public health challenge worldwide, particularly in Sub-Saharan Africa, where the disease disproportionately affects vulnerable populations. Caused by Plasmodium parasites transmitted through the bite of infected Anopheles mosquitoes, malaria accounted for an estimated 249 million cases globally in 2022, marking an increase of five million cases compared to the previous year. Uganda alone reported over 597,000 malaria cases during this period, reflecting the country’s substantial malaria burden (World Health Organization, 2023; Talapko et al., 2019). Pregnant women represent one of the most vulnerable groups to malaria infection, facing an increased risk of severe clinical symptoms and poor pregnancy outcomes. Malaria during pregnancy has been associated with a range of adverse outcomes, including miscarriage, fetal loss, preterm birth, low birth weight, and neonatal mortality (Chua et al., 2021).
